Gravitational Lens Modeling in a Citizen Science Context

### 2014-08-06 text/ms.tex
whatever
final changes for final draft
expanded fig description
corrected issue with colorspace
other minor cleanups in code
added utf8 encoding for my name

@@

@@

\author[Kueng et al]{Rafael Kueng,$^{1}$
Prasenjit Saha,$^{1}$
Elisabeth Baeten,$^{2}$
Claude Cornen,$^{2}$
\newauthor
Christine McMillan,$^{2}$
Julianne Wilcox,$^{2}$
Jonas Odermatt,$^{3}$
Jonathan Coles,$^{4}$
\newauthor
Aprajita Verma,$^{5}$
Anupreeta More,$^{6}$
Surhud More,$^{6}$
and Phil Marshall$^{7}$ \\
$^{1}$Physik-Institut, University of Zurich, 190 Winterthurerstrasse, 8057, Zurich, Switzerland\\
$^{2}$The Space Warps Collaboration\\
$^{3}$Kantonsschule Zug, L\"ussiweg 24, 6300 Zug, Switzerland\\
$^{4}$Exascale Computing Lab, Versailles Saint-Quentin-en-Yveline University, 55 Avenue des Etats Unis, 78000, Versailles, France\\
$^{5}$Oxford Astrophysics, Keble Road, Oxford OX1 3RH, UK\\
$^{6}$Kavli Institute for the Physics and Mathematics of the Universe, University of Tokyo, 5-1-5 Kashiwanoha, Kashiwa-shi 277-8583, Japan\\
$^{7}$Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, 452 Lomita Mall, Stanford, CA 94035, USA\\
}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{fig/arriv_0}
\includegraphics[width=\columnwidth]{fig/arriv_1}
\includegraphics[width=\columnwidth]{fig/arriv_2}
\caption{Perspective views and contour maps of example arrival time
  surfaces.  Contours are coloured in rainbow order (red: least delay,
  violet: highest delay).  The special contours that self-cross at
  saddle points are the basis of spaphetti diagrams.  {\em Upper
    panel:\/} No lens, hence showing the parabolic shape of the
  geometrical time delay.  The image would be at the bottom,
  coinciding with the source. {\em Middle panel:\/} A circular lensing
  mass (offset from the source) has been added, which has pushed the
  minimum to one side and introduced a maximum and saddle point, each
  corresponding to an image.  The saddle-point is characterized by a
  self-crossing ``spaghetti'' contour. {\em Lower panel:\/} An
  elongated lensing mass has been added.  There are now two minima,
  two saddle points, and a maximum, each corresponding to an image.}
\label{fig:arriv}
\end{figure}

@@

  \caption[result 6919 (ASW0002z6f)]{Another configuration of arc plus
    counter-image; here the arc is closer to the lensing galaxy than
    the counter-image. (See \secref{example_models} for details.)}
  \label{fig:6919}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\myplotswidth]{fig/006975_input}
  \includegraphics[width=\myplotswidth]{fig/006975_arr_time_ipol} \\
  \includegraphics[width=\myplotswidth]{fig/ASW000195x_006975_arriv}
  \includegraphics[width=\myplotswidth]{fig/006975_spaghetti} \\
  \includegraphics[width=\myplotswidth]{fig/ASW000195x_006975_kappa}
  \includegraphics[width=\myplotswidth]{fig/006975_mass}
  \caption[result 6975 (ASW000195x)]{A lens with unrecovered mass
    substructure. (See \secref{example_models} for details.)}
  \label{fig:6975}
\end{figure}

@@

@@

\item Figure \ref{fig:6941} shows the simplest case, with two clear
  images produced by a nearly-circular lens.  The centre of the
  lensing galaxy is a maximum, the image nearer to the galaxy is a
  saddle point, and the image further away is a minimum.  All these
  were correctly identified.  As note above, in
  \secref{SpaghettiLens}, the precise shape of the loops in the
  spaghetti diagrams is unimportant, only the implied image locations,
  parities and time-ordering matters.

@@

@@

\item Figure \ref{fig:7022} shows another model of the same system.

@@

There is no standard way of defining the Einstein radius of a general
non-circular lens.  We adopt the simple definition
\begin{equation} \label{eq:effRE}
 \langle\kappa\rangle_{\Theta_{\rm E}} = 1,
\end{equation}
that is, the effective Einstein radius \ERf is such that the
mean $\kappa$ is unity inside a circle of radius \ERf centred at
the lens centre.

@@

\begin{figure*}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/eR_5}
  \caption{Model-recovered versus actual Einstein radii \ERf[,rec] and
    \ERf[,act].  Circles denote models where the image parities and
    time-orderings were correct, with filled circles for models by an
    expert.  Crosses mark models where the time-ordering was
    incorrect.  Plus signs indicate models flagged by the modeller as
    failures.}
  \label{fig:ER_per_sim}
\end{figure*}

@@

\Figref{ER_per_sim} shows that \ERf[,rec] of the models tend to be too
high.  However, this is entirely due to the GLASS model density
profiles being too shallow, as illustrated above. We can separate out
the performance of the \spl interface and its users by comparing their
results with the  Einstein radii of \spl models made by an expert
(PS). Discounting the models which were flagged by the volunteers as
poor, the mean Einstein radius overestimate was 25\%, with a 20\%
standard deviation (shown by the grey band in \Figref{ER_per_sim}). 
The expert models show a statistically indistinguishable bias of 20\%,
with standard deviation 8\% (the blue band in \Figref{ER_per_sim}).
One source of this systematic error is that it is difficult to centre
of the lens accurately: an offset leads to a flatter mass profile for
the model compared to the simulation.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and it is also apparent that the model
  density profiles tend to be too shallow.  The likely explanation is
  that while the sims are steeply peaked at the centre, the pixellated
  mass model fixes a comparatively large area near the central at
  constant density.  One way to solve this problem would be to
  introduce smaller pixels in the central region, thus enabling a
  steeper centre. The use of simply parameterised, appropriately
  steeply-profiled models would also avoid this problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for 
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes. 
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modeling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modeling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
\end{enumerate}

>>>>

% ====================================================================
% NOTES
%
% * in case of the following warning with png files:
%     warning: pdflatex> libpng warning: iCCP: known incorrect sRGB profile
%   do:
%     find . -type f -name "*.png" -exec convert {} -strip {} \;
%
% * in case of multiple page group error:
%    find . -type f -name "*.pdf" -exec sed -i 's/\/Group.*R//g' {} \;
%    (http://tex.stackexchange.com/a/154416/37041)
%
% ====================================================================

@@

\usepackage[utf8]{inputenc}  % for my lastname :) allows umlauts in the input file without hacks
\usepackage[T1]{fontenc}     % and a font that supports umlauts for output pdf
\usepackage[english]{babel}  % just to be clear. should be the default anyways

@@

\author[Küng et al]{Rafael Küng,$^{1}$
Prasenjit Saha,$^{1}$
Elisabeth Baeten,$^{2}$
Claude Cornen,$^{2}$
\newauthor
Christine McMillan,$^{2}$
Julianne Wilcox,$^{2}$
Jonas Odermatt,$^{3}$
Jonathan Coles,$^{4}$
\newauthor
Aprajita Verma,$^{5}$
Anupreeta More,$^{6}$
Surhud More,$^{6}$
and Phil Marshall$^{7}$ \\
$^{1}$Physik-Institut, University of Zurich, 190 Winterthurerstrasse, 8057, Zurich, Switzerland\\
$^{2}$The Space Warps Collaboration\\
$^{3}$Kantonsschule Zug, L\"ussiweg 24, 6300 Zug, Switzerland\\
$^{4}$Exascale Computing Lab, Versailles Saint-Quentin-en-Yveline University, 55 Avenue des Etats Unis, 78000, Versailles, France\\
$^{5}$Oxford Astrophysics, Keble Road, Oxford OX1 3RH, UK\\
$^{6}$Kavli Institute for the Physics and Mathematics of the Universe, University of Tokyo, 5-1-5 Kashiwanoha, Kashiwa-shi 277-8583, Japan\\
$^{7}$Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, 452 Lomita Mall, Stanford, CA 94035, USA\\
}

@@

\begin{figure}
\centering
\includegraphics[width=0.95\columnwidth]{fig/arriv_0}
\includegraphics[width=0.95\columnwidth]{fig/arriv_1}
\includegraphics[width=0.95\columnwidth]{fig/arriv_2}
\caption{Perspective views and contour maps of example arrival time
  surfaces.  Contours are coloured in rainbow order (red: least delay,
  violet: highest delay).  The special contours that self-cross at
  saddle points are the basis of spaphetti diagrams.  {\em Upper
    panel:\/} No lens, hence showing the parabolic shape of the
  geometrical time delay.  The image would be at the bottom,
  coinciding with the source. {\em Middle panel:\/} A circular lensing
  mass (offset from the source) has been added, which has pushed the
  minimum to one side and introduced a maximum and saddle point, each
  corresponding to an image.  The saddle-point is characterized by a
  self-crossing ``spaghetti'' contour. {\em Lower panel:\/} An
  elongated lensing mass has been added.  There are now two minima,
  two saddle points, and a maximum, each corresponding to an image.}
\label{fig:arriv}
\end{figure}

@@

  \caption[result 6919 (ASW0002z6f)]{Another configuration of arc plus
    counter-image; here the arc is closer to the lensing galaxy than
    the counter-image. (See \secref{example_models} for details.)
    \vspace{1em}}
  \label{fig:6919}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\myplotswidth]{fig/006975_input}
  \includegraphics[width=\myplotswidth]{fig/006975_arr_time_ipol} \\
  \includegraphics[width=\myplotswidth]{fig/ASW000195x_006975_arriv}
  \includegraphics[width=\myplotswidth]{fig/006975_spaghetti} \\
  \includegraphics[width=\myplotswidth]{fig/ASW000195x_006975_kappa}
  \includegraphics[width=\myplotswidth]{fig/006975_mass}
  \caption[result 6975 (ASW000195x)]{A lens with unrecovered mass
    substructure. (See \secref{example_models} for details.)
    \vspace{1em}
    }
  \label{fig:6975}
\end{figure}

@@

The model synthetic image presented in this paper is not the original one,
but an interpolated version generated by an updated version of \spl. The image
presented during the original experiment was more rough. The two plots in
the middle showing $t(x,y)$ have uniform, but arbitrary spaced contour lines.
The $\kappa(x,y)$ plots in the bottom row show solid lines for $\kappa>1$
and dashed lines for $\kappa<1$. The spacing is logarithmic, with 10 contours
for every decade --- that means contour spacing is a quarter magnitude in 
optical terms.

@@

\item Figure \ref{fig:6941} shows the simplest case, with two clear
  images produced by a nearly-circular lens.  The center of the
  lensing galaxy is a maximum, the image nearer to the galaxy is a
  saddle point, and the image further away is a minimum.  All these
  were correctly identified.  As note above, in
  \secref{SpaghettiLens}, the precise shape of the loops in the
  spaghetti diagrams is unimportant, only the implied image locations,
  parities and time-ordering matters. $\kappa(x,y)$ shows, that
  the model has a more shallow mass distribution than the simulation.
  This is a persistent issue throughout all models and is discussed in
  \secref{tests.t2}.

@@

\item Figure \ref{fig:6937} shows a sim with substructre, like
  Figure \ref{fig:6937}. In comparison to the above, the resulting mass
  model is poor.

@@

\item Figure \ref{fig:7022} shows another model of the same system, the
  only one done by an expert in this sample. The image parities are
  correct. The elongation has the right orientation, but is too shallow.

@@

There is no standard way of defining the Einstein radius of a general
non-circular lens.  We adopt the simple definition
\begin{equation} \label{eq:effRE}
 \langle\kappa\rangle_{\Theta_{\rm E}} = 1,
\end{equation}
that is, the effective Einstein radius \ERf is such that the
mean $\kappa$ is unity inside a circle of radius \ERf centered at
the lens center.

@@

\begin{figure*}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/eR_5}
  \caption{Model-recovered versus actual Einstein radii \ERf[,rec] and
    \ERf[,act].
%    Circles denote models where the image parities and
%    time-orderings were correct, with filled circles for models by an
%    expert.  Crosses mark models where the time-ordering was
%    incorrect.
    Plus signs indicate models flagged by the modeller as
    failures by commenting negatively about it in the forum.}
  \label{fig:ER_per_sim}
\end{figure*}

@@

\Figref{ER_per_sim} shows that \ERf[,rec] of the models tend to be too
high.  However, this is entirely due to the GLASS model density
profiles being too shallow, as illustrated above. We can separate out
the performance of the \spl interface and its users by comparing their
results with the  Einstein radii of \spl models made by an expert
(PS). Discounting the models which were flagged by the volunteers as
poor, the mean Einstein radius overestimate was 25\%, with a 20\%
standard deviation (shown by the light grey band in \Figref{ER_per_sim}). 
The expert models show a statistically indistinguishable bias of 20\%,
with standard deviation 8\% (the dark grey band in \Figref{ER_per_sim}).
One source of this systematic error is that it is difficult to center
of the lens accurately: an offset leads to a flatter mass profile for
the model compared to the simulation.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and it is also apparent that the model
  density profiles tend to be too shallow.  The likely explanation is
  that while the sims are steeply peaked at the centre, the pixellated
  mass model fixes a comparatively large area near the central at
  constant density.  One way to solve this problem would be to
  introduce smaller pixels in the central region, thus enabling a
  steeper centre. The use of simply parameterised, appropriately
  steeply-profiled models would also avoid this problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for 
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes. 
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modeling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modeling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
  Better and more detailed introductory materials would also allow
  the community of modellers to grow faster and without individual
  instructions by experts or experienced volunteers.
\end{enumerate}
